---
title: "Project - NYC flights data"
output: html_notebook
---

# Project: NYC flights data

This projects aim is to better understand flight delays at the nyc airports.
You will find the data in the `nycflights13` package, remember to install
it.

```{r}
library(tidyverse)
library(nycflights13)
library(lubridate)
```

We will be using two datasets, one called `flights` with the flight information, and
one called `weather` with the weather information. You can obtain information about
the data collected by calling the R help for these datasets

```{r}
help(flights)
help(weather)
```

We can join these datasets into a single dataset containing the flights and the weather at
the time of the flight. We have also removed an outlying observation for the weather
where the windspeed was recorded above 1000mph, and removed missing observations for the
flight departure delay. We have also set delays of more than 200 minutes to 200 minutes.

```{r}
weather = weather %>% filter(wind_speed < 1000) %>% na.omit()
flights = flights %>% na.omit() %>% mutate(dep_delay = pmin(dep_delay, 200))
all_flights = inner_join(flights, weather %>% select(-year, -month, -day, -hour, - wind_gust), by=c('origin','time_hour'))
```

Although there are 300k observations in the original dataset, we will only use 30k, a tenth of the size
of the original data, so as to avoid memory problems on the computers. We will pretend that the rest
of the data was never collected.

```{r}
all_flights = all_flights %>%
  sample_n(10000) %>%
  mutate(weekend = wday(time_hour) %in% c(1, 7))
```


## First step: visualizing the data

We first wish to investigate the departure delays, and how they depend on the circumstances
of the departure
- the distribution of the departure delays (both in aggregate and per airport)
- how the wind speed affects the departure delay
- how the rain affects the departure delay
- how the month of the year affects the departure delay
- how the hour of the day affects the departure delay


In addition, compute the proportion of flights that were later than 5, 10 and 30 minutes
for each hour of the day. When are you most likely to encounter flights that are running late?

## Linear regression

I have included a fit of all the possible relevant variables for our model below.
```{r}
fit.full = lm(dep_delay ~ wind_speed + precip + month + day + hour + origin + air_time + visib + pressure + humid + dewp + distance + weekend, data=all_flights)

fit.full %>% summary()
```

From the fit above, answer the following questions:
- which of the airports has the most delays?
- are there more delays on weekdays or weekends?
- are there more delays towards the start or the end of the day?


## Model selection

In the model above, we have included all possible variables. Suppose
instead that we wish to obtain a model that only captured the important
effect. Use best subset selection along with the BIC to select the best model.
(Note: you may need to increase `nvmax` when running `regsubsets`, check the help).

Indicate which model you have selected, and produce a fit for the model.


## Logistic regression

Instead of estimating the average delay, we could also consider estimating the
probability of facing a delay of more than 5 minutes, or 15 minutes.

Fit a logistic glm for the probability of the flight being more than 5
minutes late, and another logistic glm for the probability of the flight being
more than 15 minutes late, using the same variables that you have selected
in the previous section. (Hint: you may want to create a new dataframe and add
the desired response variables).

Are the models in line with the interpretation of your linear model from
the previous section?