---
title: "Homework 4"
output: html_notebook
---

```{r}
library(tidyverse)
```


# Simulating the 2016 American elections

The presidential election in the United States involves a complex
system known as the electoral college. This makes statistical prediction
of the outcome somewhat more of a challenge. In this homework, we will
consider simulating the 2016 election on the eve of election day,
and look into the assumptions that go into predicting the outcome.


# The electoral college

The American electoral system works with a first past the post system by state,
where the party that wins the majority in a given state wins all the electors
of that state (we will ignore the peculiarities of Maine and Nebraska).
Each state is assigned a fixed number of electors based on
the state's population. We record this information in the dataframe below,
along with Trump's and Clinton's polling average on the eve of the election (we
will also ignore the third party vote).

```{r}
state.info = read_csv('../data/polling-average.csv')
```

# Simulating the election

## Question 1: Determining the winner (3 points)

Our first step is to consider how to determine the winner of the election
from the result of the vote. Suppose that we obtain a dataframe with the
percentage of people voting for Trump in each state, and the number of
electors. Write a function to determine the total number of electors for
trump, and apply this function to the polling info to determine the
number of electors trump would win if the votes corresponded to the polls.

Hint: we may write the number of electors trump obtains as a filter operation
(only picking states where Trump has the majority) followed by an aggregation
operation (summarising all electors by their sum).

Complete the code below and indicate whether Trump would have won if the
voting reflected the polling exactly:

```{r}
num.electors = function(data) {
  # complete the function
  # It should return the number of electors
}

# compute the number of electors as if
# the voting results where the poll results
num.electors(state.info)
```


## Question 2: Accounting for errors in the polls (3 points)

However, we know that the polls are not completely accurate, and that the
voting results may differ from the polling results. Let us simulate the voting
results as a random normal, with mean corresponding to the polling result and
standard deviation 2. Write a function that computes Trump's voting share in each
state.

Complete the function below to simulate the number of votes Trump would receive
in each state:

```{r}
# This function creates a new dataframe with one column being the number of
# electors in a state, and another column being the simulated vote share
# for Trump.
num.votes = function() {
  # complete the function
}
```

Below we use the function you have produced to simulate the outcome of 1000 elections,
in terms of the number electors for trump.
```{r}
replicates = bind_rows(replicate(1000, num.votes() %>% num.electors(), simplify = F))
```

Use this to compute the probability (under this model) that Trump wins the election.
Numerous newspapers and other information sources used a model similar to this one,
which reports a very low probability of winning.
```{r}
# compute the probability that Trump wins
```


## Question 3: Correlation among errors (2 points)

Let us now consider another model. Instead of modelling independent errors per
state, we will suppose that there has been a single systematic error at
the national level. Implement a function to simulate the outcome of the election
supposing that there has been a polling error nationwide, which is modelled as
a normal with mean 0 and standard deviation 2.

```{r}
num.votes.national = function() {
  # complete the function
}
```

```{r}
replicates = bind_rows(replicate(1000, num.votes.national() %>% num.electors(), simplify = F))
```

Using the simulated results, compute the probability that Trump wins the election.
Note that although the magnitude of the error is the same, 
```{r}
# compute the probability that Trump wins
```

## Question 4: Fat tails and unlikely events (2 points)

The normal distribution happens to have tails that decay extremely fast.
That is, the probability that a normal variable takes values much larger
than its mean is extremely small. In particular, normal errors can
underestimate the magnitude of errors. Consider the graph of the
upper tail probabilities below:

```{r}
y = gather(tibble(normal.tail = pnorm(seq(0, 3, 0.01), lower.tail = F),
                  t.tail = pt(seq(0, 3, 0.01), df=2, lower.tail = F),
                  x = seq(0, 3, 0.01)), dist, p, normal.tail:t.tail)

ggplot(data = y) + geom_line(aes(x = x, y = p, colour = dist))
```

We will use this property to produce a more conservative estimate of the
probability of winning. Write a function as in the previous question assuming
a single error at the national level, except model this error as twice a
t distributed random variable with degrees of freedom 2 and non-centrality
parameter 0 (hint: use the `rt` function and look at the documentation).
```{r}
num.votes.t = function() {
  # write a function as in question 4, but generate the error as twice
  # a t variable.
}
```

We generate 1000 replicates again.
```{r}
replicates = bind_rows(replicate(1000, num.votes.t() %>% num.electors(), simplify = F))
```

Use the replicates to compute the probability that Trump wins the votes
under the assumption of a t-distributed error.
```{r}
# compute the probability that Trump wins.
```

This model is now fairly similar to that obtained by 538, which used a more complex
model based on the same ideas that there might be correlation in the errors. They
gave a final estimate of about 30\% probability of winning.
