---
title: "Homework 1"
output:
  html_document: default
  html_notebook: default
---

# Homework 1

In this homework, you will be asked to analyze a dataset of the price of
diamonds. You are expected to provide both the result and the code used.
Please submit both your .Rmd file, and the knitted html or pdf document.

```{r}
library(tidyverse)
```

The dataset already exists in R and is called diamonds.
```{r}
diamonds
```
The R help describes information about the dataset. It is included below for your
reference:

- price: price in US dollars (\$326-\$18,823)
- carat: weight of the diamond (0.2-5.01)
- cut: quality of the cut (Fair, Good, Very Good, Premium, Ideal)
- color: diamond colour, from J (worst) to D (best)
- clarity: a measurement of how clear the diamond is (I1 (worst), SI1, SI2, VS1, VS2, VVS1, VVS2, IF (best))
- x: length in mm (0-10.74)
- y: width in mm (0-58.9)
- z: depth in mm (0-31.8)
- depth: total depth percentage = z / mean(x, y) = 2 * z / (x + y) (43-79)
- table: width of top of diamond relative to widest point (43-95)

## Question 1 (0.5 point)

For the following variables, indicate whether they are numerical or categorical,
and give further details about their types where appropriate: price, carat, color.

Answer below:


## Question 2 (1 point)

Compute the average price and the average weight (carat) of the diamonds sold.

Answer below:

```{r}
mean(diamonds$carat)
mean(diamonds$price)
```


## Question 3 (1 point)

We wish to better understand the distribution of the weight (carat). Plot a histogram
representing the distribution of the `carat` variable, with a bin width of 0.05.
You may note some spikes in the data. Can you try to interpret those spikes?

Answer below:

```{r}
ggplot(data=diamonds) + geom_histogram(aes(x=carat), binwidth=0.05)
```

We see some spikes around integer and half integer values. Indeed, in the process
of cutting diamonds, jewelers may prefer to target a "round" value.

## Question 4 (1.5 points)

A common hypothesis is that heavier diamonds are more expensive. Use an appropriate
visualization to confirm or invalidate this hypothesis. Explain your choice.

Answer below:

As both price and weight are continuous, a scatter plot is an appropriate
visualization. We see below that the scatter plot indeed confirms this
trend.

```{r}
ggplot(data=diamonds) + geom_point(aes(x = carat, y = price))
```

## Question 5 (1.5 points)

Diamonds are rated on a color scale from J (worst) to D (best). It is natural
to think that diamonds with a better color rating are more expensive. Use the
appropriate visualization to confirm or invalidate this observation. Was
the result what you expected?

Answer below:

We are interested in understanding how a continuous variable (price) differs
by a category (color). A boxplot is an appropriate way to visualize this
phenomenon. As we see below, it seems in fact that the price seems to
increase when the color is judged worse! This strange phenomenon
is related to Simpson's paradox, and we will attempt to explain
it in the rest of the questions.

```{r}
ggplot(data=diamonds) + geom_boxplot(aes(x = color, y=price))
```

## Question 6 (1.5 points)

Let's try to understand why the behaviour observed in question 5 does not match
our expectation. We know from question 4 that the weight influences the price
of a diamond significantly. In this question, we wish to only look at diamonds
whose weight are between 1.0 and 1.2 carats. Reproduce the visualization you
have chosen in question 5 on this reduced dataset. Comment on the interpretation.

Answer below:

We now filter down to the weight being between 1 and 1.2, and use a boxplot
again. It seems now that the worst colors have become less expensive. This
points to a confounding effect of weight on the price and color.

```{r}
ggplot(data= diamonds %>% filter(1 <= carat & carat < 1.2)) +
  geom_boxplot(aes(x=color, y=price))
```

## Question 7 (1.5 points)

It seems that taking the weight into account has somewhat resolved the paradox.
It would be helpful to understand how the weight, price and color relate. Choose an
appropriate visualization of the relationship of these three variables. Can you interpret
the reason why diamonds with worse color seem more expensive in question 5?

Answer below:

We consider adding color as a third attribute to the plot. Another option would be
to facet on the color. We see that at a given weight, the "worse" colors are
cheaper and the "better" colors are more expensive. However, the worse colors
have significantly more larger diamonds, which are just as expensive.

```{r}
ggplot(data=diamonds) + geom_point(aes(x=carat,y=price,colour=color))
```


## Question 8 (1.5 point)

Compute the average weight of the diamonds sold for each colour. Use an aggregation
operation to do so. Do the values support your explanation in question 7? Compare
those averages to the grand average you computed in question 1. Which colors tend
to be smaller, and which tend to be larger?

Answer below:

We see that the better the color is, the smaller is the average diamond. As the
weight has an important impact on price, this explains the trend that we observe
that over all the data, diamonds of worse color seem to be more expensive.
```{r}
diamonds %>% group_by(color) %>% summarise(AvgWeight = mean(carat))
```