---
title: "Introduction to R"
output: html_notebook
---

# Introduction

R is as set of software designed to facilitate most statistical tasks.
We will be using R throughout the course to both illustrate examples
and analyse data.

This type of document is called a R notebook. A R notebook allows us
to intermingle text and code freely. Currently, all this is simple text,
and is not treated in any special manner by R.

To execute some R code, we need to insert a R chunk.
You can either use the menu on the upper right, use the keyboard
shortcut *Ctrl-Alt-I*, or type the triple backtick character. 
```{r}
1 + 2
```

To execute the code, press *Ctrl-Shift-Enter*. The result is displayed
below the code chunk. R can evaluate most arithmetic expressions
(try some out!).

```{r}
1 + 4
5 * 6
```

# Basics of R code

R is a programming language, which allows us to specify complex operations
that will be executed by the computer. Let's start with some basics
of computer languages.

## Values

R can represent different types of values natively. The most important are
numbers, which we can just type in, and strings, which need to be quoted

```{r}
1 + 5
'Hello World'
```

## Variables
Variables can be thought of named containers or boxes that can store a
value. For example, let's assign the result of `1 + 4` to variable `a`, and
display its value

```{r}
a = 1 + 4
a
```
Just like in maths, a variable can take the place of any value where appropriate.
For example, let's compute 5 times the value of a

```{r}
a * 5
```

It is also possible to update the value of the variable by assigning to it again,
for example, let's put that value back into `a`
```{r}
a = a * 5
a
```
Now `a` is 5 times the value it was.

## Vectors

R is not only able to represent real values, but also vectors (i.e. list of numbers).
We can create a vector by calling the function `c`.

```{r}
c(1,2,3,4)
```

Variables can of course also contain vector values
```{r}
b <- c(1,2,3,4)
b
```

As vectors are created with `c`, remember not to call your variable `c`!
R also has a shorthand syntax for creating a vector of numbers from 1 to n,
```{r}
1:5
```

We can access individual components of the vector by using the indexing operator.
Unlike many other languages, the first component is at index 1.
```{r}
b[4]
```

R is also able to do arithmetic on vector by doing the arithmetic componentwise.
For example, multiplication between two vectors multiplies the corresponding components
```{r}
c(1,3,5) * c(2,1,5)
```

## Packages

R has a wide community of users, and those have contributed numerous packages that
are very useful and helpful in doing complex tasks. We can easily install a package
by going to Tools > Install Package, or simply calling the `install.packages` function.
Here we install the tidyverse package which we will be using throughout. You can
find documentation for most the package [here](http://tidyverse.org/).

```{r}
install.packages('tidyverse')
```

Once the package is installed, we need to load the package to make use of it.
Simply do so by calling `library`
```{r}
library(tidyverse)
```

## Data frames

R has a built-in feature to handle dataframes, although we will be using
a slight improvement from tidyverse named `tibble`. We can create a simple
dataframe like this

```{r}
df = tibble(A = c(60, 75, 150), B = c(4, 5, 6), Z = c(30, 30, 30))
df
```

We can ask for an element of the table as expected, for example consider
the element at the first row, second column
```{r}
df[1, 2]
```

We can also asks for a full row or full column at a time by index, but also
by name (not how RStudio autocomplets for you after the `$` sign).
```{r}
df[,1]
```

```{r}
df$A
```

We will usually be loading data from files on disk. R also has built in functions
to do so. For example, let's read in the MLB player data. Note that in R, the `.`
character has no significance and can be part of any variable name.

```{r}
data.mlb = read_csv('../data/socr-mlb.csv')
data.mlb
```

The functions `nrow` and `ncol` can be useful to get the number of rows and columns
of the data

```{r}
nrow(data.mlb)
ncol(data.mlb)
```

## Summary statistics in R
R has built-in functions to compute common summary statistics. The easiest to use
is simply called `summary`, and provides some summary statistics when called on a dataframe.

```{r}
summary(data.mlb)
```

We can also make use of the functions `mean`, `var`, `sd`, `median`, for example
```{r}
mean(data.mlb$Age)
var(data.mlb$Age)
sd(data.mlb$Age)
```

R also has a built-in help feature, if you're not sure of what `sd` does, we can
check the help

```{r}
help(sd)
```

## Subsetting data

We will often be interested in particular subsets of the data. THe `filter` function
from `dplyr` (part of `tidyverse`) can help us to that. For example, let's get all
players who are older than 29.

```{r}
filter(data.mlb, Age > 29)
```

We can combine logical expressions using the and combinator, &, and the or combinator, |.
For example, display players who are older than 29 and weigh more than 200 pounds.

```{r}
filter(data.mlb, Age > 29 | Weight > 200)
```

