---
title: "Statistical visualization"
output: html_notebook
---

# Why visualization?

Visualization can be a powerful tool to understand the structure of your
data. It significantly improves on the expressiveness of descriptive
summary statistics whilst still being understandable.

# Tools of visualization

Although R has an integrated plotting tools (which I will refer to as
"base plot"), one of the most compelling argument for using R is the
existence of the `ggplot2` package. It implements a unified and
systematic approach to plotting inspired by a *grammar of graphics*.

This document serves as an introduction to both the `ggplot2` package
and the basic ideas in statistical vizualizations. The notes also
provide a more theoretical description of the package.


```{r}
# As always, we use the tidyverse. ggplot2 is one part of it
library(tidyverse)
# Lubridate is a useful package to work with dates.
# Install and load it
install.packages('lubridate')
library(lubridate)
```

We will mostly be illustrating our examples using the following
two datasets. The first is a record of about 1000 MLB players.
The second is the record of the usage of the Washington D.C.
bike share program.

Remember to click on Session -> Set Working Directory -> To Source File Location
before loading the data to make sure your computer finds it!

```{r}
data.mlb <- read_csv('../data/socr-mlb.csv')
data.mlb
```
```{r}
data.bike <- read_csv('../data/bike-sharing.csv')
data.bike
```

# First steps in ggplot2 - the bar chart

Our first step in ggplot2 is the bar chart, which can be used to visualize
a single categorical variable.

A ggplot2 plot has three main components:
- the data
- the aesthetic map
- the geometry

We start the plot by calling the `ggplot` function specifying the
appropriate data that we want to display. We then select the type
of geometry we would like to use to visualize the data. This determines
the type of plot we are doing (bar, point, line, etc.). Finally,
for each graphical dimension of the chart, we specify the corresponding
variable that it should represent. This is called the aesthetic map.
In the case of the bar chart, there isn't much to specify. We tell
ggplot that we would like to see the position on the x-axis.

Note the trailing `+` sign on the first line. R is a whitespace-sensitive
language, and if we put the + on the second line, R would interpret the
first line as over, which is undesirable.
```{r}
ggplot(data=data.mlb) +
  geom_bar(aes(x = Position))
```

We may then make minor modifications such as modifying the label orientation
by adding on to the plot with the desired function
```{r}
ggplot(data=data.mlb) +
  geom_bar(aes(x = Position)) +
  theme(axis.text.x = element_text(angle=90, hjust=1)) +
  xlab('Player position') +
  ggtitle('Bar chart of count of player by position')
```

# Visualizing one continuous variable

If we wish to visualize a single continuous variable, the most commonly
used plot is the histogram. In ggplot2, this can simply be accessed
using the `geom_histogram` geometry. By default, ggplot2 plots 30 bins,
but it is often desirable to select the bin width ourselves.
```{r}
ggplot(data=data.mlb) + geom_histogram(aes(x=Height))
```

As the bin width is a property of the geometry, this is where we should specify
it. The aesthetic mapping (i.e. what goes inside `aes`) denotes the data dependent
quantities. The bin width instead has nothing to do with the observations and
is a value we picked.
```{r}
ggplot(data=data.mlb) + geom_histogram(aes(x=Height), binwidth=1)
```

An alternative to the histogram is the kernel density plot. It is similar
in concept in that it attempts to represent the distribution of the observations,
but approaches this in a more continuous fashion. It can be seen as attempting
to estimate the *density* of the variable, a notion we will cover in the
probability section of the course.
```{r}
ggplot(data=data.mlb) + geom_density(aes(x=Height))
```

# Visualizing two continuous variables

Going beyond the simple case of one variable, statistical visualizations
can also help us understand the relationship between two variables. The
visualization we use will be guided by the type of the variables we
wish to visualize.

## Scatter plots

For two numerical variables, the commonly used scatter plot is simple
way of visualizing the relationship between the two variables.
Unlike the previous visualizations that only had one dimension,
we now need to provide a map for both the `x` and `y` dimension,
hence the two arguments in `aes`.
```{r}
ggplot(data=data.mlb) + geom_point(aes(x = Height, y = Weight))
```

This allows us to easily see the positive association between the
height and weight. It also reveals that the height has only been measured
up to 0.5 inches precision, which would have been difficult to observe
otherwise.

## Line plots
Another common visualization for two numerical variables is used when
the observation are sequential. For example, whilst the baseball players
where observed in no particular order, the bikeshare program usage was
recorded in chronological order. In those cases, it may be sensible
to leverage this information in the visualization, in the form of a line
plot.

Here, we plot the bike usage data.
```{r}
ggplot(data=data.bike) +
  geom_line(aes(x=datetime, y=count))
```
However, it does not appear to be very legible, as the amount of data we have
plotted is very large. We do notice some seasonal patters with winters having
fewer users.

We may consider a smaller portion of the data, by filtering out the data to
only include the first week of June 2011 for example. This combines the same
plot as above with the filtering technique from the introductory slide.
```{r}
ggplot(data=filter(data.bike, '2011-06-01' < datetime & datetime < '2011-06-8')) +
         geom_line(aes(x=datetime, y=count))
```

Lines are appropriate when the data is collected in a fairly continuous fashion.
If there are big gaps between the data points, or if the points are fairly
discrete, a bar chart may be more appropriate. For example, suppose instead
we looked at the monthly usage numbers. In this case, the bar chart is
more appropriate as it does not imply any smoothness in the variation
and faithfully represents the fact that these are aggregates. This is
related to the notion that discretized numerical variables are categorical.
```{r}
# these lines get the aggregate counts by month
# we will learn how to do this later in the course
data.bike.monthly = data.bike %>%
  group_by(month=floor_date(datetime, unit='month')) %>%
  summarise(count=sum(count))

ggplot(data = data.bike.monthly) +
  geom_bar(aes(x=month, y=count), stat='identity')
```

## Visualization of one categorical and one numerical variable

In this section, we are interested in understanding how one numerical
variable varies across the groups identified by one categorical variable.
The most common method is the box plot, which displays a graphic representing
the summary variables for each group. This allows us to easily visualize
the differences between the groups in term of location and spread.
```{r}
ggplot(data=data.mlb) +
  geom_boxplot(aes(x=Position, y=Weight))
```

We may similarly use the same principle to fix the label positions as we had
for the bar chart.
```{r}
ggplot(data=data.mlb) +
  geom_boxplot(aes(x=Position, y=Weight)) +
  theme(axis.text.x = element_text(angle=90, hjust=1))
```

In some cases, we might still be worried that the simple descriptive statistics
do not reflect for the data for some other reason. In this case, we may choose
to display a more complex visualization for each group. The violin plot
summarises each group by using a kernel density estimate we have seen earlier,
but flipped on the side.

```{r}
ggplot(data=data.mlb) +
  geom_violin(aes(x=Position, y=Weight)) +
  theme(axis.text.x = element_text(angle=90, hjust=1))
```

## Visualizing more than two variables

In more complex datasets, we may often wish to visualize the relationship
between more than two variables at a time. However, most of the position
based visualization techniques are restricted to two dimensions. To visualize
more than two variables, there are two possibilities: we can either create
several plots side by side (so-called faceting), or map the third variable
to a characteristic that is not the x and y position.

### Faceting

The idea of faceting when plotting more than two variables is to consider
two main variables of interest, and make a plot for each group determined
by the remaining variables. For example, suppose we wished to look at the
number of bikeshare users per hour using a box plot.
```{r}
# this creates a dataset with the hour in
data.bike.hour <- data.bike
data.bike.hour$hour <- factor(hour(with_tz(data.bike.hour$datetime, tz='US/Eastern')))
data.bike.hour
```
```{r}
ggplot(data=data.bike.hour) + 
  geom_boxplot(aes(x=hour, y=count))
```

Now, as we saw previously, there are actually two main patters in bike usage,
depending on whether the day is a working day or not. A easy way to visualize
such a pattern could be to produce two plots, with one containing the data for
working days, and one for non-working days. We can accomplish this in ggplot2
by using the `facet_wrap` command, and indicating the variable we wish to facet
on.
```{r}
ggplot(data=data.bike.hour) + 
  geom_boxplot(aes(x=hour, y=count)) +
  facet_wrap(~ workingday)
```

### Other dimensions

An alternative to faceting would be to plot the data on the same plot, essentially
super-imposing the two faceted plots, but instead distinguishing them by some secondary
characteristic. For example, we could choose different colours for the working days
and non-working days. In ggplot2, we simply need to indicate an additional aesthetic
map, which links a visible characteristic (in this case the colour) to a given variable
in our data (the working day).
```{r}
data.bikeh <- data.bike
data.bikeh$hour <- factor(hour(with_tz(data.bikeh$datetime, tz='US/Eastern')))
ggplot(data=data.bikeh) + 
  geom_boxplot(aes(x=hour, y=count, colour=workingday))
```

However, we should note that as the number of categories increase, characteristics
such as colour can become more difficult to read and distinguish. For example,
let's try plotting the weight and height of MLB players, and colour-code the points
according to the player's position. However, due to the large number of different
positions, and the not too dissimilar patters, we can only gain little insight
by doing so.
```{r}
ggplot(data=data.mlb) +
  geom_point(aes(x = Height, y = Weight, colour=Position))
```

On the other hand, a facet approach allows us to better isolate some groups that
do seem to appear somewhat different from the other groups. In general, using
other characteristics such as size, colour or linetype can be useful when the
number of groups are small (< 5), but when the number of groups becomes large,
a faceting approach will usually lead to more readable visualizations.
```{r}
ggplot(data=data.mlb) +
  geom_point(aes(x = Height, y = Weight)) +
  facet_wrap(~Position)
```