---
title: "Processing data"
output: html_notebook
---

```{r}
library(tidyverse)
```


# Why process data

Although the amount of data collected has vastly increased over the past
decade, the format in which the data is collected will not always be the
most adequate for you to do your analysis. Most often, we will need to
use software tools to transform our data into a format that is adequate
for us to reason about and make use of existing software tools.


# Contingency tables
When working with categorical data, we will often want to summarise
the data into contingency tables.
```{r}
data.mlb = read_csv('../data/socr-mlb.csv')
data.mlb
```
For example, suppose that we are interested looking at how many players
of each position does each team have. This is a simple contigency table
of players and teams. To do so, we simply use the `table` function in R,
which builds a contingency table from the indicated variables.
```{r}
table(data.mlb$Team, data.mlb$Position)
```

# Grouping and aggregating

Contingency tables are a particular case of a more general idea, which is
to group the data according to some variables, and compute some aggregate of
the group. In the case of a contingency table above, the groups correspond
to a combination of team and position, and the aggregate is simply the count
of the observations.

Suppose instead, that we wish to obtain some of the statistics similar to what
we had in the boxplot. For example, let's compute the average weight of a player
of a given position. We will be using many features of
[dplyr](http://dplyr.tidyverse.org/).
```{r}
data.mlb %>%
  group_by(Position) %>%
  summarise(AvgWeight = mean(Weight))
```
This code has a couple of new features, let's walk through them together. First,
notice the `%>%` operator that we are using. This is the 'pipe' operator, and is
used to feed the data through a series of process. In this case, we process the
data in two steps. We first indicate that we would like to create groups according
to the player's position by using the `group_by` function, then for each group, we
summarise that group by taking the average weight and calling that `AvgWeight`.

We could also compute more summaries for each group, suppose for example that
we also wanted the average height. Then, we would only need to change the
`summarise` part, as we still desire the same groups.

```{r}
data.mlb %>%
  group_by(Position) %>%
  summarise(AvgWeight = mean(Weight), AvgHeight = mean(Height))
```

# Combining grouping and filtering

In addition to grouping on some variables, we will often also be required to filter
on some other variables. The `%>%` operator makes it easy to combine the filtering
operations with the grouping and aggregating. Suppose for example that we only wish
to look at players over the age of 29. We only need to add a filtering step to our
process.
```{r}
data.mlb %>%
  filter(Age > 29) %>%
  group_by(Position) %>%
  summarise(AvgWeight = mean(Weight), AvgHeight = mean(Height))
```