---
title: "Survival analysis"
output: html_notebook
---

# Survival analysis

Survival analysis deals with the problem of working with censored data.
We say data is censored when we only have partial information about the
value of the information.

```{r}
library(tidyverse)
library(survival)
library(KMsurv)
library(survminer)
library(survsim)
```

## Understanding censoring mechanism

```{r}
n = 200
data.sim = tibble(start.date = runif(n, max=100), event.time = rexp(n, 0.04)) %>%
  mutate(cens.time = pmin(event.time + start.date, 100) - start.date,
         is.event = start.date + event.time < 100)

data.sim
```

```{r}
ggplot(data = data.sim %>% sample_n(20) %>% rownames_to_column('id')) +
  geom_segment(aes(x = start.date, xend = start.date + event.time, y=id, yend=id, colour=is.event)) +
  geom_vline(xintercept = 100, linetype=2)
```

```{r}
sim.surv = with(data.sim, Surv(cens.time, is.event))
sim.fit = survfit(sim.surv ~ 1)
```

```{r}
ggsurvplot(sim.fit, risk.table = T)
```

```{r}
sim.truth = tibble(x = seq(0, 80, 0.5)) %>% mutate(s = pexp(x, 0.04, lower.tail=F))

ggplot(data = bind_rows(truth = sim.truth,
                        fit = tibble(x = sim.fit$time, s = sim.fit$surv),
                        .id='type')) +
  geom_line(aes(x = x, y = s, colour=type))
```

```{r}
data.sim.complete = data.sim %>% filter(is.event)
sim.fit.complete = tibble(x = seq(0, 80, 0.5)) %>%
  rowwise() %>%
  mutate(s = mean(x < data.sim.complete$cens.time)) %>%
  ungroup()

ggplot(data=bind_rows(complete=sim.fit.complete,
                      truth=sim.truth,
                      .id='type')) +
  geom_line(aes(x = x, y = s, colour=type))
```

## A first example

A common example arises when a study is conducted for survival times
of cancer. As a patient is diagnosed with a cancer, they enter the study
and are tracked clinically. However, not all patients enter the study at the
same time. Hence at every point in time, there are some people whose death
time is unknown.

```{r}
data(tongue)
tongue
```

We collect the survival data into observed times, and also indicating whether the observation was
censored. In R, we may collect this into a survival object, where the `+` denotes censoring.
```{r}
tongue.surv = Surv(tongue$time, tongue$delta)
tongue.surv
```

Let us first simply estimate the survival curve for the data.
```{r}
tongue.fit = survfit(tongue.surv ~ 1)
tongue.fit
```

```{r}
ggsurvplot(tongue.fit, risk.table=T)
```

Now, we may be interested in comparing the survival time of the two different groups.
We can do so by simply stratifying the two groups.

```{r}
tongue.fit2 = survfit(tongue.surv ~ factor(type), data=tongue)
ggsurvplot(tongue.fit2, risk.table=T)
```

How confident are we that the two groups are different? Can
use a log-rank test to compare the two groups.
```{r}
survdiff(tongue.surv ~ factor(type), data=tongue)
```

## Cox proportional hazards

We have collected below data on recidivism. 432 male prisones were observed
for a year after being released, and the week at which they were arrested again
(or they lapsed out of the study) was recorded, along with some other characteristics.

The following variables are also included:
- fin: whether the prisoner was given financial aid. Randomized variable (controlled by researcher)
- wexp: whether the individual had full time work experience
- mar: whether the individual is married
- paro: whether the individual was released on parole
- prio: number of prior convictions
- educ: education level code
- emp<n>: whether the individual was employed during week <n>.

```{r}
data.recidivism = read.table('../data/Rossi.txt')
data.recidivism
```

Let us use a cox-proportional hazard model to model the recidivism likelihood as a function
of time and covariates.
```{r}
rec.fit = coxph(Surv(week, arrest) ~ fin + age + race + wexp + mar + paro + prio, data=data.recidivism)
rec.fit %>% summary()
```

```{r}
rec.fit %>% survfit() %>% plot()
```

## Time dependent data in cox proportional hazard

We may sometimes be interested a variable that changes with time. For example,
in this case we may be interested in how the employment status affects the probability
of recidivism.

```{r}
unfold = function(x) {
  x = as.data.frame(x)
  covariates = x %>% select(-starts_with('emp'))
  emp = x %>% select(starts_with('emp'))
  
  covariates = bind_rows(replicate(52, covariates, simplify = F)) %>%
    rownames_to_column('week.end')
  
  covariates %>%
    mutate(week.end = as.numeric(week.end),
           week.start = week.end - 1,
           emp = factor(t(emp), levels=c('no','yes')),
           arrest = week == week.end & arrest) %>%
    drop_na()
}
```

```{r}
rec.unfolded = data.recidivism %>%
  rowwise() %>%
  do(unfold(.))
```


```{r}
rec.fitemp = coxph(Surv(week.start, week.end, arrest) ~ fin + age + race + wexp + prio + emp, data=rec.unfolded)
rec.fitemp %>% summary()
```

We see that employment is very significant. However, the correlation is difficult to
interpret. It would be better to lag the employment status by 1.

```{r}
unfold.lag = function(x, lag.amount = 1) {
  x = as.data.frame(x)
  covariates = x %>% select(-starts_with('emp'))
  emp = x %>% select(starts_with('emp'))
  
  covariates = bind_rows(replicate(52, covariates, simplify = F)) %>%
    rownames_to_column('week.end')
  
  covariates %>%
    mutate(week.end = as.numeric(week.end),
           week.start = week.end - 1,
           emp = factor(t(emp), levels=c('no','yes')),
           arrest = week == week.end & arrest) %>%
    mutate(emp = lag(emp, lag.amount)) %>%
    drop_na()
}
```

```{r}
rec.unfolded_lag = data.recidivism %>%
  rowwise() %>%
  do(unfold.lag(.))
```


```{r}
rec.fitemp_lag = coxph(Surv(week.start, week.end, arrest) ~ fin + age + race + wexp + prio + emp,
                       data=rec.unfolded_lag)
rec.fitemp_lag %>% summary()
```


