---
title: "Mediation effects"
output: html_notebook
---

```{r}
library(mediation)
library(tidyverse)
```


# Mediation effects

Even in the context of experiments, we are often interested how the dependent
variable of interest depends on the independent variable directly and indirectly.

We consider below a framing example, an experiment conducted by Brader, Valentino and
Suhay (available [here](http://onlinelibrary.wiley.com/doi/10.1111/j.1540-5907.2008.00353.x/abstract))
where participants were randomly presented
different media stories about immigration, and were asked if they agreed to
send a anti-immggration letter to their member of congress about immigration.

```{r}
data(framing)
framing
```

The researcher theorised that the willingness to send the letter would be mediated by
the subject's emotional feeling during the experiment.

## Modelling mediation

In order to perform a mediation analysis, we first model the behaviour of the mediator as
a function of the treatment and pre-treatment variables (i.e. characteristics of the
individual).

```{r}
med.fit = lm(emo ~ treat + age + educ + gender + income, data=framing)
med.fit %>% summary()
```

Now modelling the outcome, we see that the emotional level is very significant
in predicting whether the subject will be willing to send a letter to their
member of congress.

```{r}
out.fit = glm(cong_mesg ~ emo + treat + age + gender + educ + income, data=framing, family=binomial('probit'))
out.fit %>% summary()
```

## Direct and mediation effects

In a mediation analysis, we wish to separate the ADE (average direct effect), the effect only
due to the treatment (controlled by the research), and the ACME (average causal mediation effect),
the "causal" effect of the mediation on the outcome.

```{r}
med.out = mediate(med.fit, out.fit, treat='treat', mediator = 'emo')
```

```{r}
summary(med.out)
```

In this case, we see that the emotional reaction mediates about 80\% of
the observed effect, and is the only significant effect. That is, we
can explain most of the observed effect in the outcome from the
emotional impact, and we are fairly confident this is causal
despite not randomising the emotional impact directly.

## Moderation in mediation effect

We may believe in addition that the mediation effect is heterogeneous,
that is, it may vary depending on the subject. In this case, it may be
of interest to model the interaction explicitly.

```{r}
mod.med = lm(emo ~ treat * age + educ + gender + income, data=framing)
mod.med %>% summary()
```

```{r}
mod.out = glm(cong_mesg ~ emo + treat * age + emo * age + educ + gender + income,
              data=framing, family=binomial('probit'))
mod.out %>% summary()
```

We may then compute the mediation effect. Let us for example compare
the mediation effect at 20 and 60 years old.
```{r}
med.age20 = mediate(mod.med, mod.out, treat='treat', mediator='emo', covariates=list(age=20))
med.age60 = mediate(mod.med, mod.out, treat='treat', mediator='emo', covariates=list(age=60))
```

```{r}
med.age20 %>% summary()
med.age60 %>% summary()
```
