---
title: "Instrument variables"
output: html_notebook
---

```{r}
library(AER)
library(tidyverse)
```


# Instrument variables

Instrument variables allow us to mimic randomized experiments even when
we do not observe all possible confounders. They rely on introducing
some external effect that is random in some sense, and which cannot
affect the outcome directly.

## Education and Income

```{r}
data.edu = ivmodel::card.data
data.edu
```

```{r}
ggplot(data=data.edu, aes(y = lwage, x = educ)) +
  geom_point() + geom_smooth()
```

```{r}
lm(lwage ~ educ + exper + expersq + black + south, data=data.edu) %>% summary()
```

```{r}
fit.iv = ivreg(lwage ~ educ + exper + expersq + black + south | exper + expersq + black + south + nearc4, data=data.edu)
fit.iv %>% summary()
```


## Price-elasticity of demand

Let us consider the problem of estimating the price-elasticity of demand of cigarettes.
The price-elasticity of demand can be modelled as the derivative of the log-demand by
the log-price. However, in this case we are not controlling the price directly, as
it is not only affected by demand but also by supply (unobserved).

```{r}
data(CigarettesSW)
data.cig = CigarettesSW %>%
  mutate(lprice = log(price),
         lquant = log(packs),
         tdiff = taxs - tax)

data.cig
```


```{r}
lm(lquant ~ lprice, data=data.cig) %>% summary()
```

However, we may believe that taxes affect the consumption of cigarettes
only through price, which allows us to use taxes as an instrumental
variable to obtain a more causal estimate.

```{r}
ivreg(lquant ~ lprice | taxs, data=data.cig) %>% summary()
```

## Intent to treat

```{r}
data.sesame = foreign::read.dta('../data/sesame.dta')
```

In some cases, we may assign a treatment to an individual, but
have no way of enforcing whether they follow the treatment.

In this case, we may consider the assignment as the instrument variable,
and whether the treatment was actually followed as the independent
variable of intereste.

We first consider a naive estimate by looking at the how we would use
a normal regression to do so:
```{r}
lm(postlet ~ regular + prelet + factor(site) + setting, data=data.sesame) %>% summary()
```

Let us compare to a regression using instrumental variables:
```{r}
ivreg(postlet ~ regular + prelet + factor(site) + setting | encour + prelet + factor(site) + setting, data=data.sesame) %>% summary()
```

