---
title: "Communities and networks"
output: html_notebook
---

```{r}
library(igraph)
library(tidyverse)
```


# Communities and networks

To understand communities and networks, we will consider the basic mathematical
object of a graph. A graph is a list of vertices (often representing individuals,
or entities), and a list of edges between vertices (representing the relationship
between those entities).

## Example of graphs

```{r}
plot(sample_gnp(40, 0.2), vertex.label=NA)
```


```{r}
plot(sample_pa(40, power=2, directed=F), vertex.label=NA)
```

```{r}
plot(sample_sbm(40, matrix(c(0.9, 0.1, 0.1, 0.9), nrow=2), c(20, 20)), vertex.label=NA)
```

## Directed vs. undirected graphs

Depending on the relationship being modelled, we may consider to model either directed
edges or undirected edges. Undirected edges are more appropriate for reciprocal
relationships, whereas directed edges can be more appropriate for non-reciprocal
relationships.

```{r}
plot(sample_pa(30, power=1, directed=T), vertex.label=NA)
```

## Graph centrality

One measure of particular interest is measuring the importance of a vertex in the network.
This is known as centrality.

```{r}
graph.sbm = sample_sbm(25, matrix(c(0.9, 0.5, 0.5, 0.1, 0.2, 0.1, 0.1, 0.1, 0.2), nrow=3), block.sizes=c(5, 10, 10),
                      directed = T)
plot(graph.sbm, vertex.label=NA)
```

```{r}
graph.centr = alpha_centrality(graph.sbm)
g = scale_color_gradient()
centr.col = g$palette((graph.centr - min(graph.centr)) / (max(graph.centr) - min(graph.centr)))
plot(graph.pa, vertex.color=centr.col, vertex.label=NA)
```


## Community detection

We consider here a network of French political blogs (in 2006), with their affiliation classified
by a French NGO project. We have plotted the network, coloured by political affilation.

```{r}
data(fblog, package = 'sand')
fblog = igraph::upgrade_graph(fblog)
```

```{r}
plot(fblog, vertex.label=NA, vertex.color = factor(igraph::vertex_attr(fblog, 'PolParty')))
```

Suppose that we did not have the true classification, and wished to understand how
the network below is clustered:
```{r}
plot(fblog, vertex.label=NA)
```

We wish to form clusters when nodes are more tightly connected.
This problem is also called community detection. In this case,
we use an algorithm based on short random walks.

```{r}
plot(igraph::cluster_walktrap(fblog), fblog, vertex.label=NA)
```