---
title: "Common distributions"
output: html_notebook
---

```{r}
# Remember to run this first
library(tidyverse)
```


# Common distributions

As R is a statistical language, it provides numerous functions to work
with common distributions.

For each distribution, R provides for functions, which provide the following
functions:

- A function r{name of distribution}, to generate random variables according
to the distribution.
- A function d{name of distribution}, which computes the p.m.f. or p.d.f.
of the distribution (density)
- A function p{name of distribution}, which computes the cumulative distribution
function (or its complement)
- A function q{name of distribution}, which computes the quantiles of the
distribution


# Normal distribution
Let us first generate 100 normal random variables with mean 1 and variance 5
```{r}
x = rnorm(mean = 1, sd = sqrt(5), n = 100)
x
```

We can plot a histogram of this sample
```{r}
ggplot() + geom_histogram(aes(x = x), bins=10)
```

We can evaluate the density function of the normal distribution
at points from -5 to 5,
```{r}
x = seq(-5, 5, 0.1)
y = dnorm(x, mean = 0, sd = 1)

ggplot() + geom_line(aes(x = x, y = y))
```

We may also evaluate the cumulative distribution function, for example, the probability
that a standard random normal variable is less than -1
```{r}
pnorm(-1, mean = 0, sd = 1)
```
We can also evaluate the probability that a standard normal variable is greater than 3
by using the `lower.tail` argument
```{r}
pnorm(3, mean = 0, sd = 1, lower.tail=F)
```

Finally, we can evaluate the quantile function. For example, let us compute the third
quartile and 97.5th percentile of the standard normal.
```{r}
qnorm(0.75, mean = 0, sd = 1)
qnorm(0.975, mean = 0, sd = 1)
```

# Poisson distribution

We now consider the Poisson example as an example.
```{r}
x = rpois(lambda = 10, n = 100)
x
```

Plot the samples we have:
```{r}
ggplot() + geom_histogram(aes(x = x), binwidth=1)
```

Plot the p.m.f. as a bar chart
```{r}
x = 0:30
y = dpois(x, lambda=10)

ggplot() + geom_bar(aes(x = x, y = y), stat='identity')
```

Compute the probability that the Poisson is less than 5.
```{r}
ppois(5, lambda = 10)
```
Compute the third quartile of the Poisson with rate 10.
```{r}
qpois(0.75, lambda=10)
```

# Other distributions

We have the following functions:

- Binomial distribution `rbinom` (note that what we have as `n` in class
is now called size)
- Uniform distribution `runif`
- Exponential distribution `rexp`

