---
title: "R Notebook"
output: html_notebook
---

# Limit theorems

This notebook helps us visualize some of the limiting behaviour of averages.

```{r}
library(tidyverse)
```


# Law of large numbers

Let us generate random normal variables with mean 1 and variance 1.
```{r}
x = rnorm(1000, mean=1, sd=1)
```

We compute the rolling averages and plot them
```{r}
x.avg = cumsum(x) / 1:length(x)
```

```{r}
ggplot() + geom_line(aes(y=x.avg, x=1:1000)) + ylim(0, 2)
```

If the variance is greater, the convergence will be slower
```{r}
x = rnorm(1000, mean=1, sd=4)
x.avg = cumsum(x) / 1:length(x)
ggplot() + geom_line(aes(y = x.avg, x=1:length(x))) + ylim(0, 2)
```

For some pathological distributions (which we will not study), the law of large
numbers does not hold.
```{r}
x = rcauchy(1000, location=1)
x.avg = cumsum(x) / 1:length(x)
ggplot() + geom_line(aes(y = x.avg, x=1:length(x)))
```

# Central limit theorem

The central limit theorem characterise the difference of the sample mean
from the true value.
```{r}
x = rpois(n=1000 * 100, lambda = 1)
x = matrix(x, nrow=100)
```
```{r}
ggplot() + geom_histogram(aes(x=x[,2]), binwidth=1)
```

We start to see that the difference looks somewhat more symmetric and bell-shaped.
We only obtain the normal distribution in the limit.
```{r}
ggplot() + geom_histogram(aes(x = colMeans(x) - 1))
```

