---
title: "R Notebook"
output: html_notebook
---

# Sampling

Sampling is important when attempting to compare statistics across groups.
In this workbook, we take a look at what can happen when we do not do
an experiment.

# Regression to the mean

The regression to the mean phenomenon creates biases when we select
on some quantity that is related to what we wish to observe.

For example, suppose that a school wishes to implement an
after-school remedial program to help students in difficulty.

After a first exam graded out of 10, the school selects the 5 worst performing
students in a class of 20, and asks them to participate in
the program. The school then looks at the result of the second test
scores, also graded out of 10. They seem to be higher, by 1.5 point on average
- can we quantify how successful the program was?

```{r}
scores.test1 = rbinom(n = 20, size = 10, p = .5)
scores.test2 = rbinom(n = 20, size = 10, p = .5)
```

```{r}
student.remedial = order(scores.test1)[1:5]

avg.test1 = mean(scores.test1[student.remedial])
avg.test2 = mean(scores.test2[student.remedial])
```

```{r}
avg.test2 - avg.test1
```

Let us think about simulating the average score discrepancy.

```{r}
regression.bias = function() {
  scores.test1 = rbinom(n=20, size=10, p=.5)
  scores.test2 = rbinom(n=20, size=10, p=.5)
  
  student.remedial = order(scores.test1)[1:5]

  avg.test1 = mean(scores.test1[student.remedial])
  avg.test2 = mean(scores.test2[student.remedial])

  avg.test2 - avg.test1
}
```

```{r}
bias.sim = replicate(1000, regression.bias())

ggplot() + geom_density(aes(x=bias.sim))
```
```{r}
mean(bias.sim)
```

In this case, the bias is quite large, about 1.8 points (out of 10) on average.
Indeed, we are in a case where every student has the same skill, and we are
simply selecting the unlucky ones.

Instead, let us consider an example where the skill of the students vary more.

```{r}
regression.bias.skill = function(p) {
  scores.test1 = rbinom(n=20, size=10, p=p)
  scores.test2 = rbinom(n=20, size=10, p=p)
  
  student.remedial = order(scores.test1)[1:5]

  avg.test1 = mean(scores.test1[student.remedial])
  avg.test2 = mean(scores.test2[student.remedial])

  avg.test2 - avg.test1
}
```

```{r}
skill.variance = 0.01
p = rbeta(20, 1 / skill.variance, 1 / skill.variance)
```

```{r}
bias.skill.sim = replicate(1000, regression.bias.skill(p))
mean(bias.skill.sim)
```

# Size bias

Let us consider the following problem: we wish to estimate the average
class size in a school. Suppose that we were to sample some students at
random, and average the size of their classes. Is this a good way of
estimating the average class size?

Suppose that there are 200 students, split in 10 classes, of the sizes given
in the code below.
```{r}
students <- tibble(student = 1:200)
classes = c(40, 40, 20, 20, 20, 15, 15, 10, 10, 10)

students$classSize = rep(classes, classes)
```

We compute the true average class size:
```{r}
mean(classes)
```

However, suppose that we sample 10 students at random, and compute the average
class size by averaging the size of the classes students are in. We obtain
a quantity that is much larger. This is due to the fact that we are more
likely to pick a student who is in a large class.
```{r}
mean(sample_n(students, 10)$classSize)
```

